Scalable gradients of Gaussian Process models with celerite

### 2018-01-24 paper/ms.tex
words and inplace grad algorithm

Gaussian Processes \citep[GPs][]{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some example include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is of central importance and this can often be the computational bottleneck.
The details of this model are omitted here (more details can be found in
\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
is that, for a dataset with $N$ data points, every evaluation of a GP model
requires computation of the inverse\footnote{Or, more specifically, a linear
operator that can left multiply a vector or a matrix by the inverse of this
matrix.} and log-determinant of the $N \times N$ covariance matrix.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Foreman-Mackey:2017}.
The details of these models can be found elsewhere
\citep{Foreman-Mackey:2017}, but the basic idea is that for some covariance
functions, the matrix $K$ can be written as a rank-$J$ semi-separable matrix
\begin{eqnarray}\eqlabel{semi-K}
K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,V^\T) +
    \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where $\bvec{a}$ is a $N$-vector, $U$ and $V$ are $N \times J$ matrices, the
diag function creates a diagonal matrix from a vector, and the tril and triu
extract the strict lower and upper triangular matrices from a dense matrix.
For a matrix of this form, \citet{Foreman-Mackey:2017} derived an algorithm
with $\mathcal{O}(N\,J^2)$ scaling for computing the Cholesky factorization
\begin{eqnarray}
K &=& L\,\mathrm{diag}(\bvec{d})\,L^\T
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{semi-L}
L &=& I + \mathrm{tril}(U\,W^\T)
\end{eqnarray}
for some $N \times J$ matrix $W$.
The algorithm from \citet{Foreman-Mackey:2017} includes an $N-1 \times J$
``preconditioning'' matrix\footnote{This matrix is called $\phi$ by
\citet{Foreman-Mackey:2017}.} that we include with the notation $P$.
The Cholesky factorization algorithm derived by \citet{Foreman-Mackey:2017} is
given is reproduced in the following algorithm:

@@

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_factor\_grad}($U$, $P$, $\bvec{d}$, $W$,
    $S$, $\bar{\bvec{d}}$, $\bar{W}$, $\bar{S}$) \\
\hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\

\hspace*{2em}$S \gets$ \texttt{zeros}($J$, $J$) \\
    \hspace*{2em}$\bvec{w}_1 \gets \bvec{w}_1 / d_{1}$ \\
\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$S \gets \texttt{diag}(\bvec{p}_{n-1})\,[
    S + d_{n-1}\,{\bvec{w}_{n-1}}^\T\,{\bvec{w}_{n-1}}
]\,\texttt{diag}(\bvec{p}_{n-1})$ \\
\hspace*{2em}\hspace*{2em}$d_{n} \gets d_{n} - \bvec{u}_n\,S\,{\bvec{u}_n}^\T$\\
\hspace*{2em}\hspace*{2em}$\bvec{w}_n \gets \left[\bvec{w}_n -
    \bvec{u}_n\,S \right] / d_{n}$\\
    \hspace*{2em}\textbf{return} $\bvec{d}$, $W$, $S$
\end{minipage}
\medskip

>>>>

Gaussian Processes \citep[GPs][]{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some example include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is of central importance and this can often be the computational bottleneck.
The details of this model are omitted here (more details can be found in
\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
is that, for a dataset with $N$ data points, every evaluation of a GP model
requires computation of the inverse\footnote{Or, more specifically, a linear
operator that can left multiply a vector or a matrix by the inverse of this
matrix.} and log-determinant of the $N \times N$ covariance matrix.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Foreman-Mackey:2017}.
The details of these models can be found elsewhere
\citep{Foreman-Mackey:2017}, but the basic idea is that for some covariance
functions, the matrix $K$ can be written as a rank-$J$ semi-separable matrix
\begin{eqnarray}\eqlabel{semi-K}
K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,{V}^\T) +
    \mathrm{triu}({V}\,{U}^\T)
\end{eqnarray}
where $\bvec{a}$ is a $N$-vector, ${U}$ and ${V}$ are $N \times J$
matrices, the diag function creates a diagonal matrix from a vector, and the
tril and triu extract the strict lower and upper triangular matrices from a
dense matrix.
For a matrix of this form, \citet{Foreman-Mackey:2017} derived an algorithm
with $\mathcal{O}(N\,J^2)$ scaling for computing the Cholesky factorization
\begin{eqnarray}
K &=& L\,\mathrm{diag}(\bvec{d})\,L^\T
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{semi-L}
L &=& I + \mathrm{tril}(U\,W^\T)
\end{eqnarray}
for some $N \times J$ matrix $W$.
The algorithm from \citet{Foreman-Mackey:2017} includes an $N-1 \times J$
``preconditioning'' matrix\footnote{This matrix is called $\phi$ by
\citet{Foreman-Mackey:2017}.} that we include with the notation $P$.
The preconditioned variables were indicated with a tilde by
\citet{Foreman-Mackey:2017}, but to simplify the notation, we will only refer
to the pre-conditioned variables in all that follows.

The Cholesky factorization algorithm derived by \citet{Foreman-Mackey:2017} is
given is reproduced in the following algorithm:

@@

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_factor\_grad}($U$, $P$, $\bvec{d}$, $W$,
    $S$, $\bar{S}$, $\bar{\bvec{a}}$, $\bar{V}$) \\
\hspace*{2em}\textsf{\# At input $\bar{\bvec{a}} = \bar{\bvec{d}}$ and
    $\bar{V} = \bar{W}$}\\
    \hspace*{2em}$\bar{\bvec{v}}_N \gets \bar{\bvec{v}}_N / d_N$\\
\hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
\hspace*{2em}\hspace*{2em}$\bar{a}_n \gets \bar{a}_n -
    \bvec{w}_n\,{\bar{\bvec{v}}_n}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_n \gets - [\bar{\bvec{v}}_n +
    2\,\bar{a}_n\,\bvec{u}_n]\,S$\\
\hspace*{2em}\hspace*{2em}$\bar{S} \gets \bar{S} -
    {\bvec{u}_n}^\T\,[\bar{\bvec{v}}_n + \bar{a}_n\,\bvec{u}_n]$\\

\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_{n-1} \gets \mathrm{diag}(\bar{S}\,S
    \,\mathrm{diag}(\bvec{p}_{n-1})^{-1} + \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,
    S\,\bar{S})$\\
\hspace*{2em}\hspace*{2em}$\bar{S} \gets \mathrm{diag}(\bvec{p}_{n-1})\,
    \bar{S}\,\mathrm{diag}(\bvec{p}_{n-1})$\\
\hspace*{2em}\hspace*{2em}$\bar{d}_{n-1} \gets \bar{d}_{n-1} +
    \bvec{w}_{n-1}\,\bar{S}\,{\bvec{w}_{n-1}}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{v}}_{n-1} \gets \bar{\bvec{v}}_{n-1}
    / d_{n-1} + \bvec{w}_{n-1}\,[\bar{S} + \bar{S}^\T]$\\

\hspace*{2em}\hspace*{2em}$S \gets \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,S\,
    \mathrm{diag}(\bvec{p}_{n-1})^{-1}
    - d_{n-1}\,{\bvec{w}_{n-1}}^\T\,\bvec{w}_{n-1}$\\

\hspace*{2em}$\bar{a}_1 \gets \bar{a}_1 -
    \bar{\bvec{v}}_1\,{\bvec{w}_1}^\T$\\

\hspace*{2em}\textbf{return} $\bar{U}$, $\bar{P}$, $\bar{\bvec{a}}$, $\bar{V}$
\end{minipage}
\medskip
